---
title: "NOOA Storm Data Analysis - Which Events Hurt Most"
output: html_document
---
## Synopsis

This analysis explores the Storm data of NOOA and tries to address two questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing


```{r packagePrep, message=FALSE}
library(knitr)
library(dplyr)
library(lubridate)
library(ggplot2)
```

```{r setOptions, echo=FALSE}
opts_chunk$set(echo = TRUE, results = "asis")
```

The dataset for this analysis is hosted on the course website: 
[Link to Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

For this analysis, it is downloaded and saved into the subfolder **data** of my working folder. From here, this dataset is analyzed using R.

1. Load the data into a data frame and examine the structure
```{r loadData, cache=TRUE}
dfStorm <- read.csv("./data/repdata-data-StormData.csv.bz2",
                   stringsAsFactors = F)
```

2. According to the questions targeted, keep only the fields most relevant to this analyis, i.e. 

* `BGN_DATE`, the date when the event began
* `EVTYPE`, the type of natural event
* `FATALITIES`, the number of lives lost caused by the event
* `INJURIES`, the number of injuries caused by the event
* `PROPDMG`, the dollar amount of property damage
* `PROPDMGEXP`, the exponential symbol for property damage dollar amount 
* `CROPDMG`, the dollar amount of crop damage
* `CROPDMGEXP`, the exponential symbol for crop damage dollar amount 
  
```{r transformData}
dfStorm <- dfStorm %>% 
    tbl_df() %>% # tbl data frame format which works better with dplyr 
    select(BGN_DATE, EVTYPE, 
           FATALITIES, INJURIES, 
           PROPDMG, PROPDMGEXP, 
           CROPDMG, CROPDMGEXP)
           
```

3. Explore the data completeness along the time and select a representative period 
+ Convert `BGN_DATE` field into right date format in R and create

```{r convertDate}
dfStorm$BGN_DATE <- mdy_hms(dfStorm$BGN_DATE)
dfStorm$YEAR <-as.factor(year(dfStorm$BGN_DATE))
```

+ Examine the number of records per year with a plot
```{r plotbyYear}
plot(dfStorm$YEAR, 
     xlab = "Year of Event",
     ylab = "Events Recorded")

```
   
  The plot indicates that earlier years have limited events recorded while later years have more complete records.
20,000 events per year can be set as a cutoff point.  

+ Choose the years whose nubmer of recorded events is above 20,000 per year 

```{r filterbyYear}
okYear <- dfStorm %>% 
    group_by(YEAR) %>%
    summarize(N = n()) %>%
    filter(N>20000) %>%
    select(YEAR) %>%
    unlist()

dfStorm <- dfStorm %>% 
    filter(YEAR %in% okYear)
```

4. Align crop and property damages dollar amounts into USD millions based on symbols in `PROPDMGEXP` or `CROPDMGEXP` fields, i.e. 
+ 'K' meaning thousands, hence to be divided by 1000 
+ 'M' meaning millions, hence no change
+ 'B' meaning billions, hence to be multiplied by 1000
+ other symbols considered to be unclean data hence zero
```{r convertEXP}
dfStorm <- dfStorm %>%
    mutate(PROPDMG = )
```


5. Group the data by event type `EVTYPE`
dfStorm <- dfStorm %>% 
    group_by(EVTYPE)
## Results 

To simplify the measure for human population health, I make an assumption that an injury is equavalent to half a fatality. Hence the new proxy measure `HealthImpact = Fatality + 0.5 * Injury` is introduced and calculated

```{r analysis1}
dfHealth <- dfStorm %>% 
    summarize(Fatality = sum(FATALITIES), 
              Injury = sum(INJURIES),
              HealthImpact = Fatality + 0.5 * Injury) %>%
    filter(HealthImpact > 0) %>% 
    arrange(desc(HealthImpact))
```

```{r analysis2}
dfEconomy <- dfStorm %>%
    summarize(Crop = sum(CROPDMG), 
              Property = sum(PROPDMG),
              Total = Crop + Property) %>%
    filter(Total > 0) %>%
    arrange(desc(Total), desc(Crop))
```